---
title: "Installing LandGenCourse"
author: "Helene Wagner"
output: 
  html_document: 
    toc: yes
---

# 1. Detach all packages except for a few basic ones

```{r}
detachAllPackages <- function() 
{
    basic.packages <- c("package:stats", "package:graphics", 
        "package:grDevices", "package:utils", "package:datasets", 
        "package:methods", "package:base")
    package.list <- search()[ifelse(unlist(gregexpr("package:", 
        search())) == 1, TRUE, FALSE)]
    package.list <- setdiff(package.list, basic.packages)
    if (length(package.list) > 0) 
        for (package in package.list) detach(package, character.only = TRUE)
}

detachAllPackages()
```

# 2. Pre-install lots of packages from CRAN

**Note:** if asked whether you want to restart R prior to package installation as some packages are loaded, try 'no'. 

Here are the `Imports` from CRAN:

```{r message=FALSE}
install.packages(c("ade4", "adegenet", "base64enc", "BiocManager", "car", 
                  "compiler", "cowplot", "data.table", "doParallel", "dplyr", 
                  "EcoGenetics", "effsize", "formatR", "gdistance", 
                  "GeNetIt", "geosphere", "ggmap", "ggplot2", "hierfstat", 
                  "httpuv", "igraph", "landscapemetrics", "lattice", "lme4", 
                  "maps", "mapplots", "maptools", "Matrix", "microbenchmark", 
                  "miniUI", "mmod", "MuMIn", "mvtnorm", "nlme", "parallel", 
                  "pegas", "PopGenReport", "poppr", "predictmeans", "proto", 
                  "profvis", "purrr", "pwr", "RANN", "raster", "rasterVis", 
                  "readr", "RColorBrewer", "remotes", "rio", "rlang",  
                  "rmarkdown", "rstudioapi", "sampling", "seqinr", "sf", "shiny", 
                  "spacetime", "sp", "spatialEco", "spatialreg", "spdep", 
                  "spmoran", "SoDA", "stringi", "Sunder", "swirl", "swirlify",
                  "testthat", "tibble", "usdm", "vegan"))
```

And a few more (`Suggests`):

```{r}
install.packages(c("checkpoint", "feather", "fields", "knitr", "tmaptools", 
                   "ggeffects", "secr"))
```

# 3. Packages from Github

Note: you may be asked of you want to install from source. Either way seems fine. 

```{r}
remotes::install_github("krlmlr/here")
remotes::install_github("kjgilbert/QstFstComp")
remotes::install_github("dyerlab/popgraph")
remotes::install_github("dyerlab/gstudio")
```

# 4. Install `LandGenCourse`!

```{r}
remotes::install_github("hhwagner1/LandGenCourse")
```

# 5. Extra stuff 

Bioclim: these will be installed for some worked examples

```{r}
BiocManager::install("LEA")
BiocManager::install("qvalue")
```

This package can be difficult to install, it may need basic installation of gdal (outside of R), where IT support may be needed. 

```{r}
install.packages("rgdal")
```

