---
title: "Week 10: Landscape resistance"
author: "Melanie Murphy, Jeff Bowman, and Robby Marrotte"
date: "`r Sys.Date()`"
show_toc: true
output:
  knitr:::html_vignette:
    toc: yes
    fig_width: 4 
    fig_height: 3.5
vignette: >
  %\VignetteIndexEntry{Week 10: Landscape resistance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## 1. Overview of Worked Example

### a) Goals 

This worked example shows how to:

- Convert spatial data into weights ("costs") based on different weighting approaches.
- Calculate landscape resistance using least-cost and commute time (analogous to circuit theory) approaches.

### b) Data set

Raster data is essentially georeferenced matrices (i.e., matrices with geographical co-ordinates). In this case, we will use spatial data from central Idaho for Columbia spotted frogs (Murphy et al. 2010). Spatial data come from classified Landsat (NLCD), spline-based climate predictions (Rehfeldt et al. 2006),
and topographically derived variables (Moore et al. 1993, Evans 1972). Landscape variables were selected based on knowledge of species' ecology and previous research (Murphy et al. 2010, Pilliod et al. 2002, Funk et al. 2005). 
Variables are Growing season precipitation (code = gsp), Elevation relief ratio (err), Compound topographic index (cti), and Frost-free period (ffp).  

### c) Required R libraries 

All required packages should have been installed already when you installed 'LandGenCourse'.

```{r message=FALSE, warning=TRUE}
require(LandGenCourse)
require(raster)
require(gdistance)
#require(sp)
#require(rgdal)
#require(igraph)
#require(ecodist)
#require(maptools)
source(system.file("extdata", "panel.cor.r", 
                            package = "LandGenCourse"))
```

### d) Import rasters (gsp, err, cti, and ffp). An .img file is an Imagine file format (a type of raster). If you use ArcMap, you can export as files .img, a file format that is read by a wide variety of programs.

```{r}
gsp <- raster("../Dropbox/LG_lab_data/gsp.img")
err <- raster("../Dropbox/LG_lab_data/err.img")
cti <- raster("../Dropbox/LG_lab_data/cti.img")
ffp <- raster("../Dropbox/LG_lab_data/ffp.img")
```

### e) In addition to spatial data, sample locations are also needed. Sample locations are located in the data folder (RALU_UTM.csv). Read in site locations (wetlands with Columbia spotted frogs).

```{r}
sites <- read.csv("../Dropbox/LG_lab_data/RALU_UTM.csv")
```

**Question 1**: What are UTMs and why might it be important to work in UTMs (as opposed to latitude and longitude)?

### f) The spatial coordinates of the sample locations have to be identified, and the 'sites' object must be turned into a spatial points data frame.

```{r}
sites <- SpatialPointsDataFrame(coords = sites[,3:4],
                                data = sites)
```

## 2. Explore the data set. Explore these rasters by plotting them. As you go through this exercise, use the plot function at each step to make sure your outputs "make sense".

```{r}
par(mfrow=c(2,2))
plot(gsp)
plot(err)
plot(cti)
plot(ffp)
```

### a) Plot the spatial points over the ffp raster. These are the locations with Columbia spotted frog samples within the study area. 

```{r}
plot(ffp)
points(sites, pch=3)
```

## 3. Setting costs 

The next step involves creating cost categories from the continuous data. For the purposes of this exercise, a single "landscape resistance" cost surface will be created,
based on merging the 4 variables, and employing expert opinion.

Rescaling the grids will take some processor time (depending on your computer).

First, use a relative ranking of the costs of the landscape variables based on expert opinion. Higher rank indicates more resistance as follows.

Landscape resistances:

	1)err, 2) ffp, 3) gsp, 4) cti

Important considerations: (1) The transition matrix in gdistance is based on conductance and not resistance; (2) if we are going to create a single landscape resistance
(i.e., add the costs together to create one synthetic landscape variable), costs need to represent relative importance of the variables. Keep in mind there are a variety of 
approaches for creating landscape resistance values. This exercise implements one simplistic approach.

To create relative conductance values, reverse the rank order of the resistance values:

	1) cti, 2) gsp, 3) ffp, 4) err 

### a) Get all of the rasters at the same scale 

The topographic variables were calculated off a 10 m DEM. The climate variables are at a 30 m resolution. In order to calculate costs, all of the rasters need to have exactly the same resolution, dimensions, and co-ordinate locations (i.e., the cells need to match up perfectly. The method of resampling (bilinear vs nearest-neighbor) depends on the type of data. Nearest-neighbor is for categorical data whereas bilinear interpolation is for continuous data.

Resample err and cti (currently at 10 m) to the dimensions of gsp(currentyl at 30 m).

```{r}
err <- resample(err, gsp, method= "bilinear")
cti <- resample(cti, gsp, method= "bilinear")
```

### b) Calculate conductance values

To use the below tools, conductance values must be calculated. First, look at the range of the variable in a given raster. Then apply a function to get the desired
relative conductance values. Make sure to use the plot function and visually inspect your conductances surfaces.

```{r}	
err
```

class: RasterLayer
dimensions: 746, 782, 583372 (nrow, ncol, ncell)
resolution: 30,30(x,y)
extent : 677292.5, 700762.5, 4987413, 5009783 (xmin,xmax,ymin,ymax)
coord.ref.: +proj=utm+zone=11+elips=GRS80+towgs84=0,0,0,0,0,0,0+units=m+no_defs
values: [unknown pathway]\err.img
min value: 0.04129123
max value: 0.7641473

Elevation relief ratio (err) is identifying major topographic features. In this first case, the goal is for err to have the lowest conductance values compared to our other 
landscape variables. Greater err means more change in topography in a given area. So, the higher the value, the more resistance to a Columbian spotted frog. Note that the
function below gives a linear relationship between the raw value and the cost. Any type of functional relationship could be used however.


```{r}	
err.cost <- (1/err)
err.cost
```
class: RasterLayer
dimensions: 746, 782, 583372 (nrow, ncol, ncell)
resolution: 30,30(x,y)
extent : 677292.5, 700762.5, 4987413, 5009783 (xmin,xmax,ymin,ymax)
coord.ref.: +proj=utm+zone=11+elips=GRS80+towgs84=0,0,0,0,0,0,0+units=m+no_defs
values: in memory
min value: 1.294346
max value: 25.50958

```{r}	
ffp
```

class: RasterLayer
dimensions: 746, 782, 583372 (nrow, ncol, ncell)
resolution: 30,30(x,y)
extent : 677292.5, 700762.5, 4987413, 5009783 (xmin,xmax,ymin,ymax)
coord.ref.: +proj=utm+zone=11+elips=GRS80+towgs84=0,0,0,0,0,0,0+units=m+no_defs
values: [unknown pathway]\ffp.img
min value: 0
max value: 117

```{r}	
ffp.cost <- (ffp/5)
ffp.cost
```

class: RasterLayer
dimensions: 746, 782, 583372 (nrow, ncol, ncell)
resolution: 30,30(x,y)
extent : 677292.5, 700762.5, 4987413, 5009783 (xmin,xmax,ymin,ymax)
coord.ref.: +proj=utm+zone=11+elips=GRS80+towgs84=0,0,0,0,0,0,0+units=m+no_defs
values: in memory
min value: 0
max value: 23.4


```{r}	
gsp
```

class: RasterLayer
dimensions: 746, 782, 583372 (nrow, ncol, ncell)
resolution: 30,30(x,y)
extent : 677292.5, 700762.5, 4987413, 5009783 (xmin,xmax,ymin,ymax)
coord.ref.: +proj=utm+zone=11+elips=GRS80+towgs84=0,0,0,0,0,0,0+units=m+no_defs
values: [unknown pathway]\gsp.img
min value: 197
max value: 338.0697

```{r}	
gsp.cost <- (gsp-196)/15
gsp.cost
```
class: RasterLayer
dimensions: 746, 782, 583372 (nrow, ncol, ncell)
resolution: 30,30(x,y)
extent : 677292.5, 700762.5, 4987413, 5009783 (xmin,xmax,ymin,ymax)
coord.ref.: +proj=utm+zone=11+elips=GRS80+towgs84=0,0,0,0,0,0,0+units=m+no_defs
values: in memory
min value: 0.06666667
max value: 9.471311

```{r}	
cti
```

class: RasterLayer
dimensions: 746, 782, 583372 (nrow, ncol, ncell)
resolution: 30,30(x,y)
extent : 677292.5, 700762.5, 4987413, 5009783 (xmin,xmax,ymin,ymax)
coord.ref.: +proj=utm+zone=11+elips=GRS80+towgs84=0,0,0,0,0,0,0+units=m+no_defs
values: [unknown pathway]\cti.img
min value: 0.897577
max value: 25.86349

```{r}	
cti.cost <- cti/5
cti.cost
```
class: RasterLayer
dimensions: 746, 782, 583372 (nrow, ncol, ncell)
resolution: 30,30(x,y)
extent : 677292.5, 700762.5, 4987413, 5009783 (xmin,xmax,ymin,ymax)
coord.ref.: +proj=utm+zone=11+elips=GRS80+towgs84=0,0,0,0,0,0,0+units=m+no_defs
values: in memory
min value: 0.1795154
max value: 5.172697

## 4. Create a single landscape conductance raster

Testing variables independently may be more appropriate depending on the research question but fo the purposes of today's lab, a single conductance raster will be employed.

```{r}	
cost1 <- (gsp.cost + cti.cost + err.cost + ffp.cost)
cost1
```
class: RasterLayer
dimensions: 746, 782, 583372 (nrow, ncol, ncell)
resolution: 30,30(x,y)
extent : 677292.5, 700762.5, 4987413, 5009783 (xmin,xmax,ymin,ymax)
coord.ref.: +proj=utm+zone=11+elips=GRS80+towgs84=0,0,0,0,0,0,0+units=m+no_defs
values: in memory
min value: 8.569514
max value: 35.95375

**Question 2**: Plot your cost surface with your sample locations on top. What does this tell you?

## 5. Convert conductance into effective distance 

The higher the conductance, the lower the cost or resistance of a cell, and vice versa. We want to integrate conductance across cells to derive some measure of effective (or ecological) distance. 

### a) Create a transition layer

Transition layers are constructed from a raster, and take the geographic references (projection, resolution, extent) from the original raster object. It also contains a matrix of probability of movement between cells which can be interpreted as "conductance". Each cell in the matrix represents a cell in the original raster object.

The first step is to construct a transition object based on "cost1" (which is a conductance layer as calculated). This step is computationally intensive and make take a few minutes to run.

Connections can be set based on 4, 8, or 16 neighbor rules. A value of 8 connects all adjacent cells in 8 directions. 

```{r}	
tr.cost1 <- transition(cost1, transitionFunction=mean, directions=8) 
tr.cost1
```
class: RasterLayer
dimensions: 746, 782, 583372 (nrow, ncol, ncell)
resolution: 30,30(x,y)
extent : 677292.5, 700762.5, 4987413, 5009783 (xmin,xmax,ymin,ymax)
coord.ref.: +proj=utm+zone=11+elips=GRS80+towgs84=0,0,0,0,0,0,0+units=m+no_defs
values: conductance
matrix class: dsCMatrix

### b) Visually inspect the raster

```{r}	
plot(raster(tr.cost1))
```

### c) Correct for geometric distortion

Transition values are calculated based on values of adjacent cells in the cost raster. However, we used an 8 neighbor rule and the center of diagonally connected
raster cells are farther apart from each other than the orthogonally connected cells. We are using UTM co-ordinates; however, in lat-long projections, cell sizes become smaller as you move poleward.

Values of the matrix have to be corrected for the first type of distortion for our analysis (and we would need to correct for the second type of distortion if we were using lat-longs).

```{r}	
tr.cost1 <- geoCorrection(tr.cost1,type = "c",multpl=FALSE)
```

## 6. Create a cost-distance matrix

The cost-distance matrix is based on the corrected transition layer and the site locations (as a spatial points file). The distance function requires conductance values, even though distance will be 1/conductance (i.e., resistance). The cost distance is a function of the transition (tr.cost1, a transition object) and spatial locations (sites, a spatial object). This is a single least-cost path between each pair of sites.

```{r}	
cost1.dist <- costDistance(tr.cost1,sites)
plot (cost1.dist)
```

## 7. Create a cost-distance matrix based on random paths

Commute-time distance is an alternative to effective resistance from circuit theory. Commute-time is the expected time it takes for a random walk between nodes, and has been shown to correlate highly with effective resistance (Marrotte and Bowman 2017).

```{r}	
comm1.dist <- commuteDistance(x = tr.cost1, coords = sites)
plot (comm1.dist)
```

## 8. Create a distance table to compare different cost distance estimates, and then look at correlations between the methods.

```{r}	
dist_df <- data.frame("cost1.dist"=as.numeric(cost1.dist),
                      "comm1.dist"=as.numeric(comm1.dist))
```

### a) Look at correlations between the 2 different cost distances.

```{r}	
corr.LCD.comm <- cor(dist_df$cost1.dist, dist_df$comm1.dist, method = "spearman")
corr.LCD.comm

```

**Question 3**: Are the distance measures similar or different? Why? 

**Question 4**: What is the major difference? What are the implications of this difference? When might you use one or other of the methods? 

## 9. How does changing resolution change these metrics? 

Create a loop that runs through a sequential coarsening of the raster and calculate 
these metrics and compare them to the finer resolution raster metric using a Spearman correlation. The loop will take a few minutes to run.

### a) Create loop

```{r}
cor_cost <- c()
cor_comm <- c()
res_fact <- seq(2,20,2)
for(fac in res_fact){
  cost1_agg <- aggregate(cost1, fact = fac)
  tr.cost_agg <- transition(cost1_agg, transitionFunction=mean, directions=8)
  tr.cost_agg <- geoCorrection(tr.cost_agg,type = "c",multpl=FALSE)
  cost.dist_agg <- costDistance(tr.cost_agg,sites)
  comm.dist_agg <- commuteDistance(x = tr.cost_agg, coords = sites)
  cost.dist_agg <- as.numeric(cost.dist_agg)
  comm.dist_agg <- as.numeric(comm.dist_agg)
  cor_cost <- c(cor_cost,cor(dist_df$cost1.dist, cost.dist_agg, method = "spearman"))
  cor_comm <- c(cor_comm,cor(dist_df$comm1.dist, comm.dist_agg, method = "spearman"))
}
rm(i)
```

### b) Plot the results

```{r}
plot(y = cor_cost, x = res_fact, col = 2, pch = 19, 
     ylim = c(0.9,1), xlab = "Aggregation factor", ylab = "Spearman correlation")
points(y = cor_comm, x = res_fact, col = 3, pch = 19)
legend("bottomleft", legend = c("Costdist","Commdist"), pch = 19, col = 2:3)

```

**Question 5**: What effect does changing resolution have on these metrics? 


## 10. References.

Evans IS (1972) General geomorphometry, derivatives of altitude, and descriptive statistics. In: Chorley RJ (ed) Spatial analysis in geomorphology. Harper & Row, New York.

Funk WC, Blouin MS, Corn PS, Maxell BA, Pilliod DS, Amish S, Allendorf FW (2005) Population structure of Columbia spotted frogs (Rana luteiventris) is strongly affected by the landscape. Mol Ecol 14(2): 483-496

Marrotte RR, Bowman J (2017) The relationship between least-cost and resistance distance. PLOS ONE 12(3): e0174212

McRae BH, Dickson BG, Keitt TH, Shah VB (2008) Using circuit theory to model connectivity in ecology, evolution, and conservation. Ecology 89 (10):2712-2724

Moore I, Gessler P, Nielsen G, Petersen G (eds) (1993) Terrain attributes and estimation methods and scale effects. Modeling change in environmental systems. Wiley, London.

Murphy MA, Dezanni RJ, Pilliod D, Storfer A (2010) Landscape genetics of high mountain frog populations. Mol Ecol 19:3634-3649.

Pilliod DS, Peterson CR, Ritson PI (2002) Seasonal migration of Columbia spotted frogs (Rana luteiventris) among complementary resources in a high mountain basin. Can J Zool 80: 1849-1862

Rehfeldt GE, Crookston NL, Warwell MV, Evans JS (2006) Empirical analysis of plan-climate relationships for western United States. International Journal of Plan Sciences 167: 1123-1150

```{r message=FALSE, warning=TRUE, include=FALSE}
LandGenCourse::detachAllPackages()
```
