---
title: "Week 7: Spatial Linear Models"
author: "Helene Wagner"
date: "`r Sys.Date()`"
show_toc: true
output:
  knitr:::html_vignette:
    toc: yes
    fig_width: 4 
    fig_height: 3.5
vignette: >
  %\VignetteIndexEntry{Week 7: Spatial Linear Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## 1. Overview of Worked Example

### a) Goals 

This worked example shows:

- How to test regression residuals for spatial autocorrelation.
- How to fit a model with spatially autocorrelated errors (GLS)
- How to derive Moran Eigenvector Maps from sampling locations
- How to perform spatial filtering with MEM
- How to do the analogue of a partial Mantel test with 'Sunder'

### b) Data set

Here we analyze population-level data for 29 ponds from the Columbia spotted frog data (Murphy et al. 2010). 

NOTE: IMPORT DATA, ADD DESCRIPTION
- **ralu.loci**: Data frame with populations and genetic data (181 rows x 9 columns). Included in package 'LandGenCourse'. To load it, type: data(ralu.loci)

### c) Required R libraries

All required packages should have been installed already when you installed 'LandGenCourse'.

Note: the function 'library' will always load the package, even if it is already loaded, whereas 'require' will only load it if it is not yet loaded. Either will work.

```{r message=FALSE, warning=TRUE}
#require(adegenet)
#require(gstudio)
require(LandGenCourse)
#require(tibble)
#require(here)
```

## 2. Non-linear regression model (LM)

### a) Import data

```{r}
#Frogs.site <- read.csv("./data/RALU_data_div.csv", header=TRUE)

data(ralu.site, package="GeNetIt")
Frogs.site <- data.frame(ralu.site@coords, ralu.site@data)
tibble::as.tibble(Frogs.site)

SiteName <- c(Frogs.site$SiteName[c(1:3, 5:28)], NA, Frogs.site$SiteName[30:31])

ralu.loci <- read.csv("C:/Users/wagnerh1/Desktop/R_GitHub_Projects/DGS_LG_Labs/vignettes/Week7_Spatial_Linear_Models/RALU_loci_allpops.csv")

#Frogs <- data.frame(FrogID = paste(substr(ralu.loci$Pop, 1, 3),row.names(ralu.loci), sep="."),
#                    ralu.loci)
Frogs <- data.frame(ralu.loci)
Frogs$Pop <- as.character(Frogs$Pop)
Frogs$Pop[Frogs$Pop =="Doe  "] <- "Doe"
Frogs$Pop[Frogs$Pop =="GentainL"] <- "GentianL"
Frogs$Pop[Frogs$Pop =="GentainP"] <- "GentianP"
Frogs$Pop <- factor(Frogs$Pop)

Frogs$E <- as.character(Frogs$E)
Frogs$E[Frogs$E == "10:10"] <- "9:9"
Frogs$E <- factor(Frogs$E)

#Frogs$SiteName <- SiteName[as.numeric(Frogs$Pop)]
Frogs <- data.frame(SiteName = SiteName[as.numeric(Frogs$Pop)], Frogs)
  
Frogs.genind <- adegenet::df2genind(X=Frogs[,c(3:10)], sep=":", ncode=NULL, 
                                    ind.names= NULL, loc.names=NULL, 
                                    pop=Frogs$Pop, NA.char="NA", ploidy=2, 
                                    type="codom", strata=NULL, hierarchy=NULL)
Frogs.genind
Frogs.ecogen <- EcoGenetics::genind2ecogen(Frogs.genind)
row.names(Frogs) <- row.names(Frogs.ecogen@G)
Frogs.ecogen[["S"]] <- data.frame(Frogs[,1:2])

RA <- PopGenReport::allel.rich(Frogs.genind)$mean.richness




Frogs.ecopop <- EcoGenetics::ecogen2ecopop(Frogs.ecogen)

```

### b) Least squares regression model (LM)


```{r}
mod.lm <- lm(Shannon_Median ~ Area + Depth + CTI, data = Frogs.site)
mod.lm
```





```{r message=FALSE, warning=TRUE, include=FALSE}
LandGenCourse::detachAllPackages()
```
