---
title: "Week 2: Spatial Data"
author: "Helene Wagner"
date: "`r Sys.Date()`"
show_toc: true
output:
  knitr:::html_vignette:
    toc: yes
    fig_width: 4 
    fig_height: 3.5
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Week 2: Spatial Data}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
## 1. Overview of Worked Example

Author: Helene Wagner

This code builds on data and code from the 'GeNetIt' package by Jeff Evans and Melanie Murphy.

### a) Goals 

This worked example shows:

- How to import spatial coordinates and site attributes as spatially referenced data.  
- How to plot raster data in R and overlay sampling locations.
- How to calculate landscape metrics.
- How to extract landscape data at sampling locations and within a buffer around them.

Try modifying the code to import your own data!

### b) Data set

This code uses landscape data and spatial coordinates from 30 locations where Colombia spotted frogs (*Rana luteiventris*) were sampled for the full data set analyzed by Funk et al. (2005) and Murphy et al. (2010). Please see the separate introduction to the data set.

- ralu.site: SpatialPointsDataFrame object with UTM coordinates (zone 11) in slot @coords and 17 site variables in slot @data for 31 sites. The data are included in the 'GeNetIt' package, for meta data type: ?ralu.site

We will extract values at sampling point locations and within a local neighborhood (buffer) from six raster maps (see Murphy et al. 2010 for definitions), which are included with the 'GeNetIt' package as a SpatialPixelsDataFrame called 'rasters':

- cti:   Compound Topographic Index ("wetness")
- err27: Elevation Relief Ratio 
- ffp:   Frost Free Period
- gsp:   Growing Season Precipitation
- hli:   Heat Load Index
- nlcd:  USGS Landcover (categorical map)

### c) Required R libraries

```{r message=FALSE, warning=TRUE}
# require(LandGenCourse)
require(sp)
require(raster)
require(GeNetIt)
require(landscapemetrics) # for landscape metrics
require(SDMTools) # for landscape metrics
require(tibble)
require(dplyr) # should already be imported by landscapemetrics
```

Package 'tmaptools' not automatically installed with 'LandGenCourse':

```{r message=FALSE, warning=TRUE}
if(!require(tmaptools)) install.packages("tmaptools")
```

## 2. Import site data from .csv file

### a) Import data into 'SpatialPointsDataFrame'

The site data are already in a SpatialPointsDataFrame named 'ralu.site' that comes with the package 'GeNetIt'. Use 'data(ralu.site)' to load it. This will create an object 'ralu.site'. 

To demonstrate how to create a SpatialPointsDataFrame, we create a simple data frame 'Sites' with the coordinates and site data.

```{r}
data(ralu.site)
class(ralu.site)
Sites <- data.frame(ralu.site@coords, ralu.site@data)
class(Sites)
as.tibble(Sites)
```

To illustrate importing spatial data from Excel, here we export the data as a csv file, import it again as a data frame, then convert it to a SpatialPointsDataFrame. First we create a folder 'output' if it does not yet exist. 

Note: to run the code, remove all the hashtags '#" at the beginning of the lines to uncomment them. This part assumes that you have writing permission on your computer. Alternatively, try setting up your R project folder on an external drive where you have writing permission.

The second line exports the data in 'Sites' as a .csv file. The third line re-imports the .csv file to re-create data frame 'Sites'.

```{r}
#require(here)
#if(!dir.exists(paste0(here(),"/output"))) dir.create(paste0(here(),"/output"))
#write.csv(Sites, file=paste0(here(),"/output/ralu.site.csv"), 
#          quote=FALSE, row.names=FALSE)
#Sites <- read.csv(paste0(here(),"/output/ralu.site.csv"), header=TRUE)
```

The dataset 'Sites' contains two columns with spatial coordinates and 17 attribute variables. So far, R treats the spatial coordinates like any other quantitative variables. To let R know this is spatial information, we import it into a spatial object type, a 'SpatialPointsDataFrame' from the 'sp' package.

The conversion is done with the function 'coordinates', which takes a data frame and converts it to a spatial object of the same name. The code is not very intuitive. 

Note: the tilde symbol '~' (here before the first coordinate) is often used in R formulas, we will see it again later. It roughly translates to 'is modeled as a function of'.

```{r}
Sites.sp <- Sites
coordinates(Sites.sp) <- ~coords.x1+coords.x2
as.tibble(Sites.sp)
```

Now R knows these are spatial data and knows how to handle them. It does not treat the coordinates as variables anymore, hence the first column is now 'SiteName'.

### b) Add spatial reference data

Before we can combine the sampling locations with other spatial datasets, such as raster data, we need to tell R where on earth these locations are (georeferencing). This is done by specifying the 'Coordinate Reference System' (CRS) or a 'proj4' string. 

For more information on CRS, see: https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf

We know that these coordinates are UTM zone 11 (Northern hemisphere) coordinates, hence we can use a helper function to find the correct 'proj4string', using function 'get_proj4' from the 'tmaptools' package. (For the Southern hemisphere, you would add 's' after the zone: "utm11s"). Here we call the function and the package simultaneously (this is good practice, as it helps keep track of where the functions in your code come from).

```{r}
# sp::proj4string(Sites.sp) <- tmaptools::get_proj4("utm11")
```

If we had longitude and latitude coordinates, we would modify the command like this: 
proj4string(Sites.sp) <- tmaptools::get_proj4("longlat"). 

### c) Access data in 'SpatialPointsDataFrame'

As an S4 object, Sites.sp has predefined slots. These can be accessed with the @ symbol:

- @data: the attribute data
- @coords: the spatial coordinates
- @coords.nrs: the column numbers of the input data from which the coordinates were taken (filled automatically)
- @bbox: bounding box, i.e., the minimum and maximum of x and y coordinates (filled automatically)
- @proj4string: the georeferencing information

```{r}
slotNames(Sites.sp)
```

Here are the first few lines of the coordinates:

```{r}
as.tibble(Sites.sp@coords)
```

And the proj4 string:
Let's compare this to the proj4string of the original 'ralu.site' dataset

```{r}
Sites.sp@proj4string
```
The default for 'get_proj4("utm11")' results in a slightly different proj4string than the 'ralu.site' dataset. The difference is in the 'datum' argument ('WGS84' vs. 'NAD83'): 

```{r}
ralu.site@proj4string
```

Let's go with the original information and copy it:

```{r}
Sites.sp@proj4string <- ralu.site@proj4string
```

## 3. Display raster data and overlay sampling locations, extract data 

### a) Display raster data

The raster data for this project are already available in the package 'GeNetIt', under the name 'rasters', and we can load them with 'data(rasters)'. They are stored as a 'SpatialPixelsDataFrame', another S4 object type from the 'sp' package.

```{r}
data(rasters)
class(rasters)
```

However, raster data are better analyzed with the package 'raster', which has an object type 'raster'. Let's convert the data to a 'RasterStack' of 'RasterLayer' objects (i.e. a set of raster layers with the same spatial reference information).

```{r}
RasterMaps <- stack(rasters)
class(RasterMaps)
```
Printing the name of the raster stack displays a summary. A few explanations:

- **dimensions**: number of rows (nrow), number of columns (ncol), number of cells (ncell), number of layers (nlayers). So we see there are 6 layers in the raster stack.
- **resolution**: cell size is 30 m both in x and y directions (typical for Landsat-derived remote sensing data)
- **coord.ref**: projected in UTM zone 11, though the 'datum' (NAD83) is different than what we used for the sampling locations. 

```{r}
RasterMaps
```

Now we can use 'plot', which knows what to do with a raster stack.

Note: layer 'nlcd' is a categorical map of land cover types. See this week's bonus materials for how to better display a categorical map in R.

```{r fig.width=8, fig.height=5.5}
plot(RasterMaps)
```
Some layers seem to show a similar pattern. It is easy to calculate the correlation between quantitative raster layers. Here, the last layer 'ncld', is in fact categorical (land cover type), and it's correlation here is meaningless.

```{r}
layerStats(RasterMaps, 'pearson', na.rm=T)
```

### b) Change color ramp, add sampling locations 

We can specify a color ramp by setting the 'col' argument. The default is 'terrain.colors(255)'. Here we change it to 'rainbow(9)', a rainbow colorpalette with 9 color levels.

Note: To learn about options for the 'plot' function for 'raster' objects, access the help file by typing '?plot' and select 'Plot a Raster* object'.

We can add the sampling locations (if we plot only a single raster layer). Here we use 'rev' to reverse the color ramp for plotting raster layer 'ffp', and add the sites as white circles with black outlines.

```{r fig.width=4.45, fig.height=4}
par(mar=c(3,3,1,2))
plot(raster(RasterMaps, layer="ffp"), col=rev(rainbow(9)))
points(Sites.sp, pch=21, col="black", bg="white")
```

### c) Extract raster values at sampling locations

The following code adds six variables to the data slot of Sites.sp. Technically we combine the columns of the existing data frame 'Sites.sp' with the new columns in a new data frame with the same name. 

R notices the difference in projection (CRS) between the sampling point data and the rasters and takes care of it, providing just a warning. 

```{r}
Sites.sp@data <- data.frame(Sites.sp@data, extract(RasterMaps, Sites.sp))
```
What land cover type is assigned to the most sampling units? Let's tabulate them.

Note: land cover types are coded by numbers. A total of 21 sites are classified as '42'. Check here what the numbers mean: https://www.mrlc.gov/nlcd06_leg.php

```{r}
table(Sites.sp@data$nlcd)
```

## 4. Calculate landscape metrics
We are going to use the package ('landscapemetrics')[https://r-spatialecology.github.io/landscapemetrics/]. 
It is an R package for calculating landscape metrics in a tidy (for more 
information see (here)[https://www.tidyverse.org/]) workflow and is 
basically a reimplementation of ('FRAGSTATS')[https://www.umass.edu/landeco/research/fragstats/fragstats.html]. 

There are three different levels or hierarchies for landscape metrics. Firstly, 
metrics can be calculated for each single patch (neighbouring cells of the same 
class) in the landscape. Secondly, metrics can be calculated for a certain class
and lastly for the whole landscape. All these levels are implemented and easily
accessible in 'landscapemetrics'. 

All functions in landscapemetrics start with lsm_ (for landscapemetrics). 
The second part of the name specifies the level (patch - p, class - c or landscape - l). 
The last part of the function name is the abbreviation of the 
corresponding metric (e.g. enn for the euclidean nearest-neighbor distance). 

The result will be always an identical structured tibble. The first coloumn identifies 
the layer of the input (in case you use a raster stack). The second layer identifies
the level of the metric ('patch', 'class' or 'landscape'). The third coloumn identifies 
the class id. The id coloumn identifies the patch id (will be NA on class- and landscape-
level). The next coloumn is the abbreviation of the metric and lastly the actual value.

### a) Calculate patch-, class- and landscape-level landscape metrics
To calculate all available metrics on a level, you can just use the wrapper 
`calculate_metrics` and specify e.g. for the patch-level `what = "patch"`. The 
same is possible for class- and landscape-level. However, because quite a lot of 
metrics are available this is rather slow for large rasters. 

```{r}
nlcd <- raster(RasterMaps, layer="nlcd")
NLCD <- raster(RasterMaps, layer="nlcd")

# calculate all patch-level metrics using wrapper
nlcd.patch <- landscapemetrics::calculate_metrics(nlcd, what = "patch")
# nlcd.class <- landscapemetrics::calculate_metrics(nlcd, what = "class")
# nlcd.landscape <- landscapemetrics::calculate_metrics(nlcd, what = "landscape")

nlcd.patch
# nlcd.class
# nlcd.landscape

# show abbreviation of all calculated metrics
unique(nlcd.patch$metric)
# unique(nlcd.class$metric) 
# unique(nlcd.landscape$metric) 
```

Alternatively, you can either specify the desired metrics by using the `what` 
argument or just rowbind the results of the single `lsm_` functions 
(be aware that the order might be slightly different). Because the tibble is 
always structured identical, you can easily mix different levels.

```{r}
# Use the wrapper to calculate some metrics
nlcd.subset.a <- landscapemetrics::calculate_metrics(nlcd, 
                                                     what = c("lsm_c_pladj", 
                                                              "lsm_l_pr", 
                                                              "lsm_l_shdi"))

# Rowbind results for some metrics
nlcd.subset.b <- rbind(
  landscapemetrics::lsm_c_pladj(nlcd), 
  landscapemetrics::lsm_l_pr(nlcd),
  landscapemetrics::lsm_l_shdi(nlcd)
)

nlcd.subset.a
nlcd.subset.b
```

Because the metric names are only abbreviated, there is also a way to include the 
full name in the results. For the wrapper, just set `full_name = TRUE`. For the rowbinded
tibble you can use the provided tibble `lsm_abbreviations_names` that comes with 
the package and use e.g. `left_join()`. 

```{r}
nlcd.subset.a.full <- landscapemetrics::calculate_metrics(nlcd, 
                                                     what = c("lsm_c_pladj", 
                                                              "lsm_l_pr", 
                                                              "lsm_l_shdi"), 
                                                     full_name = TRUE)

nlcd.subset.b.full <- dplyr::left_join(x = nlcd.subset.b, 
                                       y = lsm_abbreviations_names, 
                                       by = "metric")

nlcd.subset.b.full
``` 

# b)
To only get the results for class 42 (forest), you can just filter the tibble. 

```{r}
# class == 42 for forest
forest.patch.metrics <- dplyr::filter(nlcd.patch, class == 42)
```

All functions make heavy use of connected components labeling to delineate patches
(neighouring cells of the same class). To get all patches of one class you can 
just use `get_patches()`. You can choose between 8-neighbor rule or 4-neighbor rule.

```{r}
# connected components labeling of landscape
cc.nlcd <- landscapemetrics::get_patches(nlcd, directions = 8)

# Show name of each class
sapply(cc.nlcd, function(x) names(x)) 

# The first list entry is class forest
cc.forest <- cc.nlcd[[1]]
```

The result will be a list containing a raster for each class present in the input
raster.

To plot the patches you can use the `show_patches` function.

```{r}
# Add code here! 
```

Let's add forest patch size to the 'Sites.sp' data. First we need to get the patch ID at each sampling location, then its size.

```{r}
# We are working on a build-in function for this at the moment
sites_forest <- tibble::as.tibble(raster::extract(x = cc.forest, y = Sites.sp, df = TRUE))

forest.patch.area <- dplyr::filter(nlcd.patch, metric == "area", 
                                               class == 42, 
                                               id %in% sites_forest$Class_42)

sites_forest <- dplyr::left_join(x = sites_forest, y = forest.patch.area,
                                 by = c("Class_42" = "id"))

Sites.sp@data$ForestPatchSize <- sites_forest$value
Sites.sp@data$ForestPatchSize[is.na(Sites.sp@data$ForestPatchSize)] <- 0 
Sites.sp@data$ForestPatchSize
```

Plot a bubble map of forest patch size at each sampling location.

```{r fig.width=4.45, fig.height=4}
par(mar=c(3,3,1,2))
bubble(Sites.sp, "ForestPatchSize", fill=FALSE, key.entries=as.numeric(names(table(Sites.sp@data$ForestPatchSize))))
```

## 5. Extract landscape metrics at sampling locations.

### a) Calculate class-level metrics in buffer around sampling locations

First we define the buffer radius (in meters) and cell size:

```{r}
Radius <- 500    # Define buffer radius
Cellsize <- 30   # Indicate cell size in meters
```

Then we create a loop through all sampling locations (all rows of the site data set), calculating class-level metrics for each one within its buffer (see video for further explanations).

```{r}
Sites.class <- list()
class.ID <- levels(ratify(NLCD))[[1]]

for(i in 1:nrow(Sites.sp@data))
{
  # Identify all cells that lie within buffer around site i:
  Buffer.cells <- extract(NLCD, Sites.sp[i,], cellnumbers=TRUE, 
                          buffer=Radius)[[1]][,1]
  
  # Copy land cover map and delete all values outside of buffer:
  Buffer.nlcd <- NLCD
  values(Buffer.nlcd)[-Buffer.cells] <- NA
  
  # Calculate class-level metrics for cells within buffer:
  Result <- ClassStat(Buffer.nlcd,cellsize=Cellsize)
  
  # Merge Results table with 'class.ID' to ensure that all cover types
  # are listed for all sites, even if they are not present in buffer,
  # write results into ith element of list 'Sites.class':
  Sites.class[[i]] <- merge(class.ID, Result, all=TRUE, by.x="ID", by.y="class")
}
# Add labels for list elements
names(Sites.class) <- Sites.sp@data$SiteName
```

```{r}
as.tibble(Sites.class[[2]])
```
### b) Extract landscape metric of choice for a single cover type (as vector)

Now we can extract any variable of interest for any cover type of interest. Here we'll extract the percentage of evergreen forest within a 500 m radius around each site. 

```{r}
# Extract one variable, 'prop.landscape', for one cover type 42 (Evergreen Forest)
# (this returns a vector with a single value for each site)

PercentForest500 <- rep(NA, length(Sites.class))  # Create empty results vector
for(i in 1:length(Sites.class))
{
  # For site i, select row with cover type '42' and column 'prop.landscape':
  PercentForest500[i] <- Sites.class[[i]][class.ID$ID==42, "prop.landscape"]
}

# If there are any sites with no forest in buffer, set value to 0:
PercentForest500[is.na(PercentForest500)] <- 0

# Print results:
PercentForest500
```
### c) Extract landscape metric of choice for all cover types (as data frame)

To extract the landscape metric 'prop.landscape' for all cover types as a data.frame (one column per cover type), use this code. 

We'll define column names combining 'Prop' for 'proportion of landscape', '500' to indicate the 500 m buffer radius, and the ID of each cover type.

```{r}
# Create empty matrix for storing results:
Prop.landscape <- matrix(data=NA, nrow=length(Sites.class), ncol=length(class.ID$ID))

# Create row and column names:
dimnames(Prop.landscape) <- list(names(Sites.class),
                                 paste("Prop.500", class.ID$ID, sep="."))

# For each site i, extract "prop.landscape" for all cover types
# and write results into row i of Prop.landscape:
for(i in 1:length(Sites.class))
{
  Prop.landscape[i,] <- Sites.class[[i]][,"prop.landscape"]
}

# Set any missing values to 0:
Prop.landscape[is.na(Prop.landscape)] <- 0

# Convert matrix to data frame:
Prop.landscape <- as.data.frame(Prop.landscape)
as.tibble(Prop.landscape)
```

The percent cover of all cover types should add up to 100% (i.e., 1) for each site. We can check this with the function 'apply'. The argument 'MARGIN' specifies whether we want to apply the function FUN to each row (MARGIN=1) or each column (MARGIN=2). 

Note: This function expects the object X to be a matrix or array - taking a row total only makes sense if all columns contain the same type of data, in the same units. It still does the calculation here even though we just converted 'Prop.landscape' to a data frame. Always double check whether what you ask R to calculate makes sense.

```{r}
apply(X=Prop.landscape, MARGIN=1, FUN=sum)
```

### d) Extract all landscape metrics for a single cover type (as data frame)

To extract all landscape metrics for a single cover type, we need to modify the code like this. Here we add the class ID '42' to all variable names to indicate that these are quantified for cover type '42' (evergreen forest)

```{r}
# Create empty matrix for storing results:
Forest.class <- matrix(data=NA, nrow=length(Sites.class), 
                       ncol=ncol(Sites.class[[1]]))

# Create row and column names:
dimnames(Forest.class) <- list(names(Sites.class),
                                 paste("42",names(Sites.class[[1]]), sep="."))

# For each site i, extract all landscape metrics for cover type 42
# and write results into row i of Forest.class:
for(i in 1:length(Sites.class))
{
  Forest.class[i,] <- unlist(Sites.class[[i]][class.ID$ID==42,])
}

# Convert matrix to data frame:
Forest.class <- as.data.frame(Forest.class)
as.tibble(Forest.class)
```


### e) Append to site data set

```{r}
Sites.sp@data <- data.frame(Sites.sp@data, Prop.landscape,
                                Forest.class) 
```

Done!

Note: check this week's bonus material if you want to see how to use the new 'sf' library for spatial data, and how to export the site data to an shapefile that you can import into a GIS.

The following code detaches all packages except for some basic ones:

```{r message=FALSE, warning=TRUE, include=FALSE}
# LandGenCourse::detachAllPackages()
```
