---
title: "Week 2: Spatial Data"
author: "Helene Wagner"
date: "`r Sys.Date()`"
show_toc: true
output:
  knitr:::html_vignette:
    toc: yes
    fig_width: 4 
    fig_height: 3.5
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Week 2: Spatial Data}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
## 1. Overview of Worked Example

Author: Helene Wagner

This code builds on data and code from the 'GeNetIt' package by Jeff Evans and Melanie Murphy.

### a) Goals 

This worked example shows:

- How to import spatial coordinates and site attributes as spatially referenced data.  
- How to plot raster data in R and overlay sampling locations.
- How to calculate landscape metrics.
- How to extract landscape data at sampling locations and within a buffer around them.

Try modifying the code to import your own data!

### b) Data set

This code uses landscape data and spatial coordinates from 30 locations where Colombia spotted frogs (*Rana luteiventris*) were sampled for the full data set analyzed by Funk et al. (2005) and Murphy et al. (2010). Please see the separate introduction to the data set.

- ralu.site: SpatialPointsDataFrame object with UTM coordinates (zone 11) in slot @coords and 17 site variables in slot @data for 31 sites. The data are included in the 'GeNetIt' package, for meta data type: ?ralu.site

We will extract values at sampling point locations and within a local neighborhood (buffer) from six raster maps (see Murphy et al. 2010 for definitions), which are included with the 'GeNetIt' package as a SpatialPixelsDataFrame called 'rasters':

- cti:   Compound Topographic Index ("wetness")
- err27: Elevation Relief Ratio 
- ffp:   Frost Free Period
- gsp:   Growing Season Precipitation
- hli:   Heat Load Index
- nlcd:  USGS Landcover (categorical map)

### c) Required R libraries

```{r message=FALSE, warning=TRUE}
# require(LandGenCourse)
require(sp)
require(raster)
require(GeNetIt)
require(landscapemetrics) # for landscape metrics
require(SDMTools) # for landscape metrics
require(tibble)
require(dplyr) # should already be imported by landscapemetrics
```

Package 'tmaptools' not automatically installed with 'LandGenCourse':

```{r message=FALSE, warning=TRUE}
if(!require(tmaptools)) install.packages("tmaptools")
```

You will need the development version of `landscapemetrics v.0.2`.

```{r}
# CRAN update on its way at the moment
devtools::install_github("r-spatialecology/landscapemetrics", ref = "stable")
```

## 2. Import site data from .csv file

### a) Import data into 'SpatialPointsDataFrame'

The site data are already in a SpatialPointsDataFrame named 'ralu.site' that comes with the package 'GeNetIt'. Use 'data(ralu.site)' to load it. This will create an object 'ralu.site'. 

To demonstrate how to create a SpatialPointsDataFrame, we create a simple data frame 'Sites' with the coordinates and site data.

```{r}
data(ralu.site)
class(ralu.site)
Sites <- data.frame(ralu.site@coords, ralu.site@data)
class(Sites)
as.tibble(Sites)
```

To illustrate importing spatial data from Excel, here we export the data as a csv file, import it again as a data frame, then convert it to a SpatialPointsDataFrame. First we create a folder 'output' if it does not yet exist. 

Note: to run the code, remove all the hashtags '#" at the beginning of the lines to uncomment them. This part assumes that you have writing permission on your computer. Alternatively, try setting up your R project folder on an external drive where you have writing permission.

The second line exports the data in 'Sites' as a .csv file. The third line re-imports the .csv file to re-create data frame 'Sites'.

```{r}
#require(here)
#if(!dir.exists(paste0(here(),"/output"))) dir.create(paste0(here(),"/output"))
#write.csv(Sites, file=paste0(here(),"/output/ralu.site.csv"), 
#          quote=FALSE, row.names=FALSE)
#Sites <- read.csv(paste0(here(),"/output/ralu.site.csv"), header=TRUE)
```

The dataset 'Sites' contains two columns with spatial coordinates and 17 attribute variables. So far, R treats the spatial coordinates like any other quantitative variables. To let R know this is spatial information, we import it into a spatial object type, a 'SpatialPointsDataFrame' from the 'sp' package.

The conversion is done with the function 'coordinates', which takes a data frame and converts it to a spatial object of the same name. The code is not very intuitive. 

Note: the tilde symbol '~' (here before the first coordinate) is often used in R formulas, we will see it again later. It roughly translates to 'is modeled as a function of'.

```{r}
Sites.sp <- Sites
coordinates(Sites.sp) <- ~coords.x1+coords.x2
as.tibble(Sites.sp)
```

Now R knows these are spatial data and knows how to handle them. It does not treat the coordinates as variables anymore, hence the first column is now 'SiteName'.

### b) Add spatial reference data

Before we can combine the sampling locations with other spatial datasets, such as raster data, we need to tell R where on earth these locations are (georeferencing). This is done by specifying the 'Coordinate Reference System' (CRS) or a 'proj4' string. 

For more information on CRS, see: https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf

We know that these coordinates are UTM zone 11 (Northern hemisphere) coordinates, hence we can use a helper function to find the correct 'proj4string', using function 'get_proj4' from the 'tmaptools' package. (For the Southern hemisphere, you would add 's' after the zone: "utm11s"). Here we call the function and the package simultaneously (this is good practice, as it helps keep track of where the functions in your code come from).

```{r}
# sp::proj4string(Sites.sp) <- tmaptools::get_proj4("utm11")
```

If we had longitude and latitude coordinates, we would modify the command like this: 
proj4string(Sites.sp) <- tmaptools::get_proj4("longlat"). 

### c) Access data in 'SpatialPointsDataFrame'

As an S4 object, Sites.sp has predefined slots. These can be accessed with the @ symbol:

- @data: the attribute data
- @coords: the spatial coordinates
- @coords.nrs: the column numbers of the input data from which the coordinates were taken (filled automatically)
- @bbox: bounding box, i.e., the minimum and maximum of x and y coordinates (filled automatically)
- @proj4string: the georeferencing information

```{r}
slotNames(Sites.sp)
```

Here are the first few lines of the coordinates:

```{r}
as.tibble(Sites.sp@coords)
```

And the proj4 string:
Let's compare this to the proj4string of the original 'ralu.site' dataset

```{r}
Sites.sp@proj4string
```
The default for 'get_proj4("utm11")' results in a slightly different proj4string than the 'ralu.site' dataset. The difference is in the 'datum' argument ('WGS84' vs. 'NAD83'): 

```{r}
ralu.site@proj4string
```

Let's go with the original information and copy it:

```{r}
Sites.sp@proj4string <- ralu.site@proj4string
```

## 3. Display raster data and overlay sampling locations, extract data 

### a) Display raster data

The raster data for this project are already available in the package 'GeNetIt', under the name 'rasters', and we can load them with 'data(rasters)'. They are stored as a 'SpatialPixelsDataFrame', another S4 object type from the 'sp' package.

```{r}
data(rasters)
class(rasters)
```

However, raster data are better analyzed with the package 'raster', which has an object type 'raster'. Let's convert the data to a 'RasterStack' of 'RasterLayer' objects (i.e. a set of raster layers with the same spatial reference information).

```{r}
RasterMaps <- stack(rasters)
class(RasterMaps)
```
Printing the name of the raster stack displays a summary. A few explanations:

- **dimensions**: number of rows (nrow), number of columns (ncol), number of cells (ncell), number of layers (nlayers). So we see there are 6 layers in the raster stack.
- **resolution**: cell size is 30 m both in x and y directions (typical for Landsat-derived remote sensing data)
- **coord.ref**: projected in UTM zone 11, though the 'datum' (NAD83) is different than what we used for the sampling locations. 

```{r}
RasterMaps
```

Now we can use 'plot', which knows what to do with a raster stack.

Note: layer 'nlcd' is a categorical map of land cover types. See this week's bonus materials for how to better display a categorical map in R.

```{r fig.width=8, fig.height=5.5}
plot(RasterMaps)
```
Some layers seem to show a similar pattern. It is easy to calculate the correlation between quantitative raster layers. Here, the last layer 'ncld', is in fact categorical (land cover type), and it's correlation here is meaningless.

```{r}
layerStats(RasterMaps, 'pearson', na.rm=T)
```

### b) Change color ramp, add sampling locations 

We can specify a color ramp by setting the 'col' argument. The default is 'terrain.colors(255)'. Here we change it to 'rainbow(9)', a rainbow colorpalette with 9 color levels.

Note: To learn about options for the 'plot' function for 'raster' objects, access the help file by typing '?plot' and select 'Plot a Raster* object'.

We can add the sampling locations (if we plot only a single raster layer). Here we use 'rev' to reverse the color ramp for plotting raster layer 'ffp', and add the sites as white circles with black outlines.

```{r fig.width=4.45, fig.height=4}
par(mar=c(3,3,1,2))
plot(raster(RasterMaps, layer="ffp"), col=rev(rainbow(9)))
points(Sites.sp, pch=21, col="black", bg="white")
```

### c) Extract raster values at sampling locations

The following code adds six variables to the data slot of Sites.sp. Technically we combine the columns of the existing data frame 'Sites.sp' with the new columns in a new data frame with the same name. 

R notices the difference in projection (CRS) between the sampling point data and the rasters and takes care of it, providing just a warning. 

```{r}
Sites.sp@data <- data.frame(Sites.sp@data, extract(RasterMaps, Sites.sp))
```
What land cover type is assigned to the most sampling units? Let's tabulate them.

Note: land cover types are coded by numbers. A total of 21 sites are classified as '42'. Check here what the numbers mean: https://www.mrlc.gov/nlcd06_leg.php

```{r}
table(Sites.sp@data$nlcd)
```

## 4. Calculate landscape metrics

We are going to use the package [`landscapemetrics`](https://r-spatialecology.github.io/landscapemetrics/). It is an `R` package for calculating landscape metrics in a tidy workflow (for more information see [here](https://www.tidyverse.org/]) and is basically a reimplementation of ['FRAGSTATS'](https://www.umass.edu/landeco/research/fragstats/fragstats.html). 

There are three different levels for landscape metrics. Firstly, metrics can be calculated for each single patch (neighbouring cells of the same class). Secondly, metrics can be calculated for a certain class and lastly for the whole landscape. All these levels are implemented and easily accessible in `landscapemetrics`. 

All functions in landscapemetrics start with `lsm_` (for landscapemetrics). The second part of the name specifies the level (patch - `p`, class - `c` or landscape - `l`). The last part of the function name is the abbreviation of the corresponding metric (e.g. `enn` for the Euclidean nearest-neighbor distance). For all available metrics and more information about them see either the helpfile(s) or [here](https://r-spatialecology.github.io/landscapemetrics/reference/index.html).

The result of all landscape metrics functions is always an identical structured tibble. The first coloumn is the layer id (only interesting for e.g. a `RasterStack`). The second coloumn specifies the level ('patch', 'class' or 'landscape'). The third coloumn is the class id (`NA` on landscape level). The third coloumn is the patch id (`NA` on class- and landscape level). The fourth coloumn is the abbreviation of the metric and the last coloumn is the value.

### a) Calculate patch-, class- and landscape level landscape metrics 

To calculate all available metrics on a level, you can just use the wrapper `calculate_lsm` and specify e.g. for the patch-level `what = "patch"`. The same is possible for class- and landscape level. However, because quite a lot of metrics are available, this is rather slow for large rasters. 

```{r}
nlcd <- raster(RasterMaps, layer="nlcd")

# calculate all patch-level metrics using wrapper
nlcd.patch <- landscapemetrics::calculate_lsm(nlcd, what = "patch")
# nlcd.class <- landscapemetrics::calculate_metrics(nlcd, what = "class")
# nlcd.landscape <- landscapemetrics::calculate_metrics(nlcd, what = "landscape")

nlcd.patch
# nlcd.class
# nlcd.landscape

# show abbreviation of all calculated metrics
unique(nlcd.patch$metric)
# unique(nlcd.class$metric) 
# unique(nlcd.landscape$metric) 
```

Alternatively, you can either specify the desired metrics by using the `what` argument or  rowbind the results of the single `lsm_` functions (be aware that the order might be slightly different). Because the tibble is always structured identical, you can easily mix different levels.

```{r}
# use the wrapper to calculate some metrics
nlcd.subset.a <- landscapemetrics::calculate_lsm(nlcd, 
                                                 what = c("lsm_c_pladj", 
                                                          "lsm_l_pr", 
                                                          "lsm_l_shdi"))

# rowbind results for some metrics
nlcd.subset.b <- rbind(
  landscapemetrics::lsm_c_pladj(nlcd), 
  landscapemetrics::lsm_l_pr(nlcd),
  landscapemetrics::lsm_l_shdi(nlcd)
)

nlcd.subset.a
nlcd.subset.b
```

Because the metric names are only abbreviated, there is also a way to include the full name in the results. For the wrapper, just set `full_name = TRUE`. For the rowbinded tibble, you can use the provided tibble `lsm_abbreviations_names` that comes with the package and use e.g. `left_join()`. 

```{r}
# add full metrics name to result
nlcd.subset.a.full <- landscapemetrics::calculate_lsm(nlcd, 
                                                      what = c("lsm_c_pladj", 
                                                               "lsm_l_pr", 
                                                               "lsm_l_shdi"), 
                                                      full_name = TRUE)

# add full metrics name to result using left_join
# nlcd.subset.b.full <- dplyr::left_join(x = nlcd.subset.b, 
#                                        y = lsm_abbreviations_names, 
#                                        by = "metric")

nlcd.subset.a.full
# nlcd.subset.b.full
``` 

# b) Calculate patch-level landscape metrics for 'Evergreen Forest'

To only get the results for class 42 (forest), you can just `filter()` the `tibble`. 

```{r}
# class == 42 for forest
forest.patch.metrics <- dplyr::filter(nlcd.patch, class == 42)
```

All functions make heavy use of connected components labeling to delineate patches (neighouring cells of the same class). To get all patches of every class you can just use `get_patches()`. To only get a certain class, just specify the `what` argument. You can choose between 8-neighbor rule or 4-neighbor rule.

```{r}
# connected components labeling of landscape
cc.nlcd <- landscapemetrics::get_patches(nlcd, directions = 8)

# show name of each class
sapply(cc.nlcd, function(x) names(x)) 

# the first list entry is class forest
cc.forest <- cc.nlcd[[4]]
# cc.forest <- landscapemetrics::get_patches(nlcd, what = 42) # watch out: result is list with one entry

```

To plot the patches you can use the `show_patches` function.

```{r}
  # show_patches(nlcd, what = "all", labels = FALSE)
  show_patches(nlcd, what = 42, labels = FALSE)
```

It is also possible to visualize only the core area of each patch. The core area are all cells that are further away from the edge of each patch than a specified edge depth (e.g. 5 cells). 

```{r}
  show_cores(nlcd, what = 42, edge_depth = 5,  labels = FALSE)
```

Let's add forest patch size to the 'Sites.sp' data. To extract landscape metrics of a patch in which sample points are located in, use `extract_lsm()`. Which metrics are extracted can be specified by the `what` argument (similar to `calculate_lsm()`). However, only patch level metrics are available.

```{r}
# extract patch area of all classes
patch_size_sp <- extract_lsm(landscape = nlcd, y = Sites.sp, what = "lsm_p_area")

# because we are only interested in the forest patch size, we set all area of class != 42 to 0
patch_size_sp_forest <- dplyr::mutate(patch_size_sp, 
                                      value = dplyr::case_when(class == 42 ~ value, 
                                                               class != 42 ~ 0))
# add data to sp object
Sites.sp@data$ForestPatchSize <- patch_size_sp_forest$value
Sites.sp@data$ForestPatchSize
```

Plot a bubble map of forest patch size at each sampling location.

```{r fig.width=4.45, fig.height=4}
par(mar=c(3,3,1,2))
bubble(Sites.sp, "ForestPatchSize", fill=FALSE, key.entries=as.numeric(names(table(Sites.sp@data$ForestPatchSize))))
```

## 5. Sample landscape metrics in sampling plots

### a) Calculate metrics in buffer around sampling locations

`landscapemetrics` has a build-in function to sample metrics in a buffer around sample locations. You can choose the shape of the buffer (either a circle or a square) and of course which metrics to sample (similar to `calculate_lsm()`).

```{r}
# sample some metrics within buffer around sample location 
nlcd.sampled <- landscapemetrics::sample_lsm(landscape = nlcd, 
                                                 what = c("lsm_l_ta", 
                                                          "lsm_c_np",
                                                          "lsm_c_pland", 
                                                          "lsm_c_ai"),
                                                 shape = "square",
                                                 points = Sites.sp, 
                                                 size = 500)

nlcd.sampled
```

The `tibble` now contains two additional columns. Firstly, the `plot_id` (in the same order as the input points) and secondly, the `percantge_inside` of the sample plot within the landscape. In cases in which the sample location is on the edge of the landscape, the sample plot could only be partly within the landscape. The value can also deviate from 100 % because the sample locations are not necessarily in the cell center and the actually clipped cells lead to a slightly smaller or larger sample area. A circle sample shape even increases this effect.

It is also possible to get the clippings of the sample plots as a `RasterLayer`. Therefore, just set `return_plots = TRUE`.

```{r}
# sample some metrics within buffer around sample location and returning sample
# plots as raster
nlcd.sampled.plots <- landscapemetrics::sample_lsm(landscape = nlcd, 
                                                       what = c("lsm_l_ta",
                                                                "lsm_c_np",
                                                                "lsm_c_pland",
                                                                "lsm_c_ai"),
                                                       shape = "square",
                                                       points = Sites.sp, 
                                                       size = 500, 
                                                       return_plots = TRUE)

nlcd.sampled.plots
```

The result will be a nested `tibble` containing the `plot_id`, the metrics (as a `list`) and a `RasterLayer` with the clipped sample plot (as a `list`).

```{r, fig.width=8, fig.height=5.5}
# show results for first four plots
nlcd.sampled.plots$metrics[1:4]
nlcd.sampled.plots$raster_sample_plots[1:4]

par(mfrow = c(2,2))
  plot(nlcd.sampled.plots$raster_sample_plots[[1]], 
       main = paste(Sites.sp$SiteName[1]), 
       col = rev(rainbow(9)))
  plot(nlcd.sampled.plots$raster_sample_plots[[2]],
       main = paste(Sites.sp$SiteName[2]),
       col=rev(rainbow(9)))
  plot(nlcd.sampled.plots$raster_sample_plots[[3]],
        main = paste(Sites.sp$SiteName[3]), 
       col=rev(rainbow(9)))
  plot(nlcd.sampled.plots$raster_sample_plots[[4]],
        main = paste(Sites.sp$SiteName[4]), 
       col=rev(rainbow(9)))
par(mfrow = c(1,1))
```

### b) Extract landscape metric of choice for a single cover type (as vector)

To extract a metrics you can just `filter()` the resulting `tibble` and pull the `value` coloumn.
 
```{r}
# filter the results for class == 42 (forest) and pland (percentage) and pull
# results as vector
percentage.forest.500 <- dplyr::pull(dplyr::filter(nlcd.sampled, 
                                                   class == 42, 
                                                   metric == "pland"), value)

percentage.forest.500
```

### c) Extract landscape metric of choice for all cover types (as data frame). 

To extract the landscape metric 'prop.landscape' for all cover types as a `tibble`, just `filter()` the `tibble` again, but only use the `metric` as filter. 

```{r}
# filter for percentage of landscape
percentage.forest.500.df <- dplyr::filter(nlcd.sampled,
                                          metric == "pland")

percentage.forest.500.df
```

The percent cover of all cover types should add up to ~ 100% (i.e., 1) for each site. We can check this with the function `summarize()`. First, we need to group the data using the `plot_id` and than sum all percentages.

```{r}
# group by plot_id and sum all percentages
pland.sum <- dplyr::summarize(dplyr::group_by(percentage.forest.500.df, by = plot_id), 
                 sum_pland = sum(value))

pland.sum
```

### d) Extract all landscape metrics for a single cover type (as data frame)

Just `filter()` for `class == 42` and add the sites names as coloumn to the resulting `tibble`.

```{r}
# filter for class == 42 (forest)
forest.500.df <- dplyr::filter(nlcd.sampled,
                               class == 42)

# add site names to result
forest.500.df$site_name <- Sites.sp$SiteName

forest.500.df
```

Done!

Note: check this week's bonus material if you want to see how to use the new 'sf' library for spatial data, and how to export the site data to an shapefile that you can import into a GIS.

The following code detaches all packages except for some basic ones:

```{r message=FALSE, warning=TRUE, include=FALSE}
# LandGenCourse::detachAllPackages()
```
