---
title: "Solutions to R Exercises (Weeks 1 - 8)"
author: "Helene Wagner"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
---

# Week 1 

### a) Load packages

```{r}
library(gstudio)
library(tibble)
library(adegenet)
library(here)
```

### b) View data file

```{r}
# Copy to downloads folder (as needed):
file.copy(system.file("extdata", "pulsatilla_genotypes.csv", package = "LandGenCourse"),
          paste0(here(), "/downloads/pulsatilla_genotypes.csv"), overwrite=FALSE)

# Import from downloads folder:
Pulsatilla <- read.csv(paste0(here(), "/downloads/pulsatilla_genotypes.csv"), header=TRUE)
#View(Pulsatilla)
```

### c) Import data into `gstudio`

Notes: 

- Need to load library `gstudio` or use `gstudio::read_population`
- Argument `type` should be set to 'column'.
- Specify correct columns for locus data

```{r}
Pulsatilla.gstudio <- read_population(path=system.file("extdata",
                           "pulsatilla_genotypes.csv", 
                           package = "LandGenCourse"), 
                   type="column", locus.columns=c(6:19), 
                   phased=FALSE, sep=",", header=TRUE)
```

### d) Check imported data

Note: `str` may create an error message "Error: $ operator is invalid for atomic vectors" - there seems to be some interference with other packages, depending on what files are open. It seems to run without error during `knit`, though. May use `as_tibble` instead.

```{r}
str(Pulsatilla.gstudio)
# Alternative
as_tibble(Pulsatilla.gstudio) 
```

### e) Check variable types

Note: to create bulleted list, there must be an empty line before the list and one after the list.

- **ID**: integer
- **OffID**: integer
- **Population**: character
- **Coordinates**: numeric (`dbl` means numeric with double precision)
- **Loci**: locus (secondary type: character) 

### f) Convert to genind object

```{r}
Pulsatilla.genind2 <- adegenet::df2genind(X=Pulsatilla.gstudio[,c(6:12)], sep=":", ncode=NULL,   
                          ind.names=Pulsatilla.gstudio$ID, loc.names=NULL, 
                          pop=Pulsatilla.gstudio$Population, NA.char="", ploidy=2, 
                          type="codom", strata=NULL, hierarchy=NULL)
Pulsatilla.genind2
```
Note: the last row in the output ("Optional content") should list @pop and indicate that group size ranges between 55 - 128 genotyped individuals. 

### Answer

There are between 55 - 128 genotyped individuals per site ("group"). 



# Week 2 

### a) Load packages

```{r}
library(gstudio)
library(dplyr)
library(tibble)
library(sp)
```

### b) Import data

```{r}
Pulsatilla.gstudio <- read_population(path=system.file("extdata",
                           "pulsatilla_genotypes.csv", 
                           package = "LandGenCourse"), 
                   type="column", locus.columns=c(6:19), 
                   phased=FALSE, sep=",", header=TRUE)
str(Pulsatilla.gstudio)
```

### c) Summarize by site

```{r}
Pulsatilla <- Pulsatilla.gstudio %>% group_by(Population) %>% summarize(nIndiv = n())
```

### d) Add mean coordinates

Note: important to specify site indicator `Population`.

```{r}
Pulsatilla <- Pulsatilla.gstudio %>% group_by(Population) %>% summarize(nIndiv = n(), meanX = mean(X), meanY = mean(Y))
```
 
### e) Convert to spatial object

Note: important to use the variable names created above in d.

```{r}
coordinates(Pulsatilla) <- ~ meanX + meanY
```

### f) Specify known projection
 
```{r}
proj4string(Pulsatilla) <- CRS("+init=epsg:31468")
```

### g) Transform projection

```{r}
Pulsatilla.longlat <- sp::spTransform(Pulsatilla, 
                      CRSobj = tmaptools::get_proj4("longlat")$proj4string)
Pulsatilla.longlat@coords
```

### h) Create bubble plot

Note: important to use the variable name created above in c/d.
```{r}
bubble(Pulsatilla.longlat, "nIndiv", fill = FALSE)
```

### i) Save R object

```{r}
saveRDS(Pulsatilla.longlat, file = paste0(here::here(), "/output/Pulsatilla.longlat.rds"))
```

### Answer

The sites are located in Southern Germany. To check, enter the following in Google maps: 
`48.99875, 11.06123` and zoom out. 

Note: must specify latitude first, then longitude. Must use coordinates in "lonlat" projection.

# Week 3 

### a) Load packages

```{r}
library(gstudio)
library(dplyr)
library(adegenet)
```

### b) Import data

```{r}
library(gstudio)
Pulsatilla.gstudio <- read_population(path=system.file("extdata",
                           "pulsatilla_genotypes.csv", 
                           package = "LandGenCourse"), 
                   type="column", locus.columns=c(6:19), 
                   phased=FALSE, sep=",", header=TRUE)
```

### c) Count genotyped individuals

```{r}
# Overall sample size:
nrow(Pulsatilla.gstudio)

# Number of adults:
nrow(Pulsatilla.gstudio[Pulsatilla.gstudio$OffID == 0,])

# Alternative with 'filter':
Pulsatilla.gstudio %>% filter(OffID == 0) %>% nrow()
```

### d) Drop offspring from dataset

```{r}
# By indexing:
Pulsatilla.adults <- Pulsatilla.gstudio[Pulsatilla.gstudio$OffID == 0,]
nrow(Pulsatilla.adults)

# Alternative with 'filter':
Pulsatilla.adults <- Pulsatilla.gstudio %>% filter(OffID == 0)
nrow(Pulsatilla.adults)
```

### e) Split dataset by site

```{r}
Adults.by.site <- split(Pulsatilla.adults, Pulsatilla.adults$Population)
length(Adults.by.site)
```

### f) Count adults per site with `sapply`

```{r}
sapply(Adults.by.site, nrow)
```

### g) Convert to genind object

```{r}
Adults.genind <- adegenet::df2genind(X=Pulsatilla.adults[,c(6:12)], sep=":", ncode=NULL, 
                          ind.names= Pulsatilla.adults$ID, loc.names=NULL, 
                          pop=Pulsatilla.adults$Population, NA.char="", ploidy=2, 
                          type="codom", strata=NULL, hierarchy=NULL)

Adults.genind
```

### h) Check polymorphism

```{r}
summary(Adults.genind)
```

### i) Test for HWE by site and locus

Note: while a few values are <0.05, there is not systematic pattern, i.e. no locus seems to be out of HWE across many site, and vice versa.

```{r}
# Chi-squared test: p-value
HWE.test <- data.frame(sapply(seppop(Adults.genind), 
                              function(ls) pegas::hw.test(ls, B=0)[,3]))
HWE.test.chisq <- t(data.matrix(HWE.test))
{cat("Chi-squared test (p-values):", "\n")
round(HWE.test.chisq,3)}
```

### k) Calculate Hexp and Hobs by site

```{r}
Hexp <- t(sapply(seppop(Adults.genind), function(ls) summary(ls)$Hexp))
Hobs <- t(sapply(seppop(Adults.genind), function(ls) summary(ls)$Hobs))
Hexp.pop <- apply(Hexp, 1, mean)
Hobs.pop <- apply(Hobs, 1, mean)

H.pop <- data.frame(Pop = names(Hobs.pop),
                              Hobs = Hobs.pop,
                              Hexp = Hexp.pop)
H.pop
```

### l) Save results as R object

```{r}
saveRDS(H.pop, file = paste0(here::here(), "/output/H.pop.rds"))
```

### Answer

Site G05a had the lowest expected heterozygosity.


# Week 4 

# Week 5 

# Week 6 

# Week 7 

# Week 8 