---
title: "Solutions to R Exercises (Weeks 1 - 8)"
author: "Helene Wagner"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
---

# Week 1 

### a) Load packages

```{r}
library(gstudio)
library(tibble)
library(adegenet)
library(here)
```

### b) View data file

```{r}
# Copy to downloads folder (as needed):
file.copy(system.file("extdata", "pulsatilla_genotypes.csv", package = "LandGenCourse"),
          paste0(here(), "/downloads/pulsatilla_genotypes.csv"), overwrite=FALSE)

# Import from downloads folder:
Pulsatilla <- read.csv(paste0(here(), "/downloads/pulsatilla_genotypes.csv"), header=TRUE)
#View(Pulsatilla)
```

### c) Import data into `gstudio`

Notes: 

- Need to load library `gstudio` or use `gstudio::read_population`
- Argument `type` should be set to 'column'.
- Specify correct columns for locus data

```{r}
Pulsatilla.gstudio <- read_population(path=system.file("extdata",
                           "pulsatilla_genotypes.csv", 
                           package = "LandGenCourse"), 
                   type="column", locus.columns=c(6:19), 
                   phased=FALSE, sep=",", header=TRUE)
```

### d) Check imported data

Note: `str` may create an error message "Error: $ operator is invalid for atomic vectors" - there seems to be some interference with other packages, depending on what files are open. It seems to run without error during `knit`, though. May use `as_tibble` instead.

```{r}
#str(Pulsatilla.gstudio)
# Alternative
as_tibble(Pulsatilla.gstudio) 
```

### e) Check variable types

Note: to create bulleted list, there must be an empty line before the list and one after the list.

- **ID**: integer
- **OffID**: integer
- **Population**: character
- **Coordinates**: numeric (`dbl` means numeric with double precision)
- **Loci**: locus (secondary type: character) 

### f) Convert to genind object

```{r}
Pulsatilla.genind2 <- adegenet::df2genind(X=Pulsatilla.gstudio[,c(6:12)], sep=":", ncode=NULL,   
                          ind.names=Pulsatilla.gstudio$ID, loc.names=NULL, 
                          pop=Pulsatilla.gstudio$Population, NA.char="", ploidy=2, 
                          type="codom", strata=NULL, hierarchy=NULL)
Pulsatilla.genind2
```
Note: the last row in the output ("Optional content") should list @pop and indicate that group size ranges between 55 - 128 genotyped individuals. 

### Answer

There are between 55 - 128 genotyped individuals per site ("group"). 



# Week 2 

### a) Load packages

```{r}
library(gstudio)
library(dplyr)
library(tibble)
library(sp)
```

### b) Import data

```{r}
Pulsatilla.gstudio <- read_population(path=system.file("extdata",
                           "pulsatilla_genotypes.csv", 
                           package = "LandGenCourse"), 
                   type="column", locus.columns=c(6:19), 
                   phased=FALSE, sep=",", header=TRUE)
#str(Pulsatilla.gstudio)
as_tibble(Pulsatilla.gstudio)
```

### c) Summarize by site

```{r}
Pulsatilla <- Pulsatilla.gstudio %>% group_by(Population) %>% summarize(nIndiv = n())
```

### d) Add mean coordinates

Note: important to specify site indicator `Population`.

```{r}
Pulsatilla <- Pulsatilla.gstudio %>% group_by(Population) %>% summarize(nIndiv = n(), meanX = mean(X), meanY = mean(Y))
```
 
### e) Convert to spatial object

Note: important to use the variable names created above in d.

```{r}
coordinates(Pulsatilla) <- ~ meanX + meanY
```

### f) Specify known projection
 
```{r}
proj4string(Pulsatilla) <- CRS("+init=epsg:31468")
```

### g) Transform projection

```{r}
Pulsatilla.longlat <- sp::spTransform(Pulsatilla, 
                      CRSobj = tmaptools::get_proj4("longlat")$proj4string)
Pulsatilla.longlat@coords
```

### h) Create bubble plot

Note: important to use the variable name created above in c/d.
```{r}
bubble(Pulsatilla.longlat, "nIndiv", fill = FALSE)
```

### i) Save R object

```{r}
saveRDS(Pulsatilla.longlat, file = paste0(here::here(), "/output/Pulsatilla.longlat.rds"))
```

### Answer

The sites are located in Southern Germany. To check, enter the following in Google maps: 
`48.99875, 11.06123` and zoom out. 

Note: must specify latitude first, then longitude. Must use coordinates in "lonlat" projection.

# Week 3 

### a) Load packages

```{r}
library(gstudio)
library(dplyr)
library(adegenet)
```

### b) Import data

```{r}
library(gstudio)
Pulsatilla.gstudio <- read_population(path=system.file("extdata",
                           "pulsatilla_genotypes.csv", 
                           package = "LandGenCourse"), 
                   type="column", locus.columns=c(6:19), 
                   phased=FALSE, sep=",", header=TRUE)
```

### c) Count individuals

```{r}
# Overall sample size:
nrow(Pulsatilla.gstudio)

# Number of adults:
nrow(Pulsatilla.gstudio[Pulsatilla.gstudio$OffID == 0,])

# Alternative with 'filter':
Pulsatilla.gstudio %>% filter(OffID == 0) %>% nrow()
```

### d) Drop offspring 

```{r}
# By indexing:
Pulsatilla.adults <- Pulsatilla.gstudio[Pulsatilla.gstudio$OffID == 0,]
nrow(Pulsatilla.adults)

# Alternative with 'filter':
Pulsatilla.adults <- Pulsatilla.gstudio %>% filter(OffID == 0)
nrow(Pulsatilla.adults)
```

### e) Split dataset by site

```{r}
Adults.by.site <- split(Pulsatilla.adults, Pulsatilla.adults$Population)
length(Adults.by.site)
```

### f) Count adults per site 

```{r}
sapply(Adults.by.site, nrow)
```

### g) Convert to genind object

```{r}
Adults.genind <- adegenet::df2genind(X=Pulsatilla.adults[,c(6:12)], sep=":", ncode=NULL, 
                          ind.names= Pulsatilla.adults$ID, loc.names=NULL, 
                          pop=Pulsatilla.adults$Population, NA.char="", ploidy=2, 
                          type="codom", strata=NULL, hierarchy=NULL)

Adults.genind
```

### h) Check polymorphism

```{r}
summary(Adults.genind)
```

### i) Test HWE by site & locus

Note: while a few values are <0.05, there is not systematic pattern, i.e. no locus seems to be out of HWE across many site, and vice versa.

```{r}
# Chi-squared test: p-value
HWE.test <- data.frame(sapply(seppop(Adults.genind), 
                              function(ls) pegas::hw.test(ls, B=0)[,3]))
HWE.test.chisq <- t(data.matrix(HWE.test))
{cat("Chi-squared test (p-values):", "\n")
round(HWE.test.chisq,3)}
```

### k) Calculate Hexp, Hobs by site

```{r}
Hexp <- t(sapply(seppop(Adults.genind), function(ls) summary(ls)$Hexp))
Hobs <- t(sapply(seppop(Adults.genind), function(ls) summary(ls)$Hobs))
Hexp.pop <- apply(Hexp, 1, mean)
Hobs.pop <- apply(Hobs, 1, mean)

H.pop <- data.frame(Pop = names(Hobs.pop),
                              Hobs = Hobs.pop,
                              Hexp = Hexp.pop)
H.pop
```

### l) Save results as R object

```{r}
saveRDS(H.pop, file = paste0(here::here(), "/output/H.pop.rds"))
```

### Answer

Site G05a had the lowest expected heterozygosity.


# Week 4 


### a) Load packages

```{r}
library(dplyr)
library(ggplot2)
```

### b) Import saved objects 

```{r}
Pulsatilla.longlat <- readRDS(paste0(here::here(), "/output/Pulsatilla.longlat.rds"))
H.pop <- readRDS(paste0(here::here(), "/output/H.pop.rds"))
```

### c) Plot sites on map

Plot the residuals on top of the map 'myMap':

Basic map with points:

```{r}
myMap <- ggmap::qmplot(meanX, meanY,  data = as.data.frame(Pulsatilla.longlat), 
                       source = "stamen", maptype = "toner-lite")
```

Modify code from 3.c to add labels and plot the map:

```{r fig.height=5, fig.width=7, message=FALSE}
myMap + ggplot2::geom_label(data = as.data.frame(Pulsatilla.longlat),
                   mapping = ggplot2::aes(meanX, meanY, label = Population),
                   size = 2.5, label.padding = unit(0.12, "lines"),
                   col = "black", vjust = 0, nudge_x = 0, nudge_y = -0.005)
```

### d) Combine data

Note: it is important here to work with the @data slot: Pulsatilla.longlat@data.

Check structure of datasets (could also use `head` or `tibble::as_tibble` or simply print them).

```{r}
str(Pulsatilla.longlat@data)
```
```{r}
str(H.pop)
```

Change variable type from factor to character: 

```{r}
H.pop$Pop <- as.character(H.pop$Pop)
```

Add data from `H.pop` to `Pulsatilla.longlat@data`. Here it is important to indicate the ID variable names ("Population" and "Pop") correctly, and to use the @data slot. 

```{r}
Pulsatilla.longlat@data <- dplyr::left_join(Pulsatilla.longlat@data, H.pop, by=c("Population" = "Pop"))
```

Note: if the variable types are not the same for the two ID variables, you'll get a warning: "Column `Population`/`Pop` joining character vector and factor, coercing into character vector". This is only a warning, not an error message, hence R will still perform the join. 

### e) Scatterplot with line

Either with base R...

```{r}
plot(Hexp ~ nIndiv, data=Pulsatilla.longlat, xlim=c(50, 130))
abline(lm(Hexp ~ nIndiv, data=Pulsatilla.longlat))
text(Pulsatilla.longlat@data$nIndiv, Pulsatilla.longlat@data$Hexp,
     labels=Pulsatilla.longlat@data$Population, cex= 0.8, pos=4)
```

Or with ggplot:

```{r}
ggplot(Pulsatilla.longlat@data, aes(x=nIndiv, y=Hexp, label=Population)) + 
  geom_point() + 
  geom_smooth(method = lm, se = TRUE) +
  geom_text(nudge_y=0.012, size=4, check_overlap=TRUE)
```

### f) Regression analysis

Fit the model. 

Note: here you can use either `data=Pulsatilla.longlat` or `data=Pulsatilla.longlat@data`.

```{r}
mod <- lm(Hexp ~ nIndiv, data=Pulsatilla.longlat)
summary(mod)
```

The slope is not significant. Is the model valid? Checking residual plots:

```{r fig.height=7, fig.width=8}
par(mfrow=c(2,2))
plot(mod, labels.id = Pulsatilla.longlat@data$Population)
par(mfrow=c(1,1))
```

### Answers

There is one influential point in the regression analysis:

- Which site is it? - Site A25. 
- Where is it located (north / east / south / west)? - In the (South-)West.
- What makes it an influential point (large residual, leverage, or both)? - Leverage (outlier along x: very large census population size 'nIndiv')
- What would happen to the regression line and the R-squared if this point was omitted? - The line would become much steeper, and the Rsquared would be much smaller. 


# Week 5 

a) **Load packages**: You may want to load the packages `dplyr`, `EcoGenetics` and `adegenet`. Alternatively, you can use `::` to call functions from packages.

### a) Load packages

```{r}
library(adegenet)
library(dplyr)
library(EcoGenetics)
```

### b) Import, extract A25 adults

```{r}
Pulsatilla.gstudio <- gstudio::read_population(path=system.file("extdata",
                            "pulsatilla_genotypes.csv", 
                            package = "LandGenCourse"), 
                    type="column", locus.columns=c(6:19), 
                    phased=FALSE, sep=",", header=TRUE)

```

Inspect a few rows:

```{r}
as_tibble(Pulsatilla.gstudio)
```

Filter adults (OffID == 0) to drop seeds, filter site A03:

```{r}
Adults.A25.gstudio <- Pulsatilla.gstudio %>% filter(OffID == 0, Population == "A25")
```

### c) Plot locations of individuals 

```{r}
plot(Adults.A25.gstudio$X, Adults.A25.gstudio$Y, asp=1)
```

### d) Convert to ecogen and genind

Convert to `ecogen`:

```{r}
Adults.A25.ecogen <- EcoGenetics::gstudio2ecogen(Adults.A25.gstudio, ID = "ID", 
                            lat = "X", lon = "Y")
```

Convert from `ecogen` to `genind`:

```{r}
Adults.A25.genind <- EcoGenetics::ecogen2genind(Adults.A25.ecogen)
```
  
### e) Genetic & Euclidean distances

- Use individual-level function `propShared` for a `genind` object. 
- Subtract from 1 to convert to distance measure.
- Check class and convert to `dist` object.

```{r}
Dgen <- 1 - adegenet::propShared(Adults.A25.genind)
class(Dgen)
Dgen <- as.dist(Dgen)
class(Dgen)
```

Calculate Euclidean distance from coordinates:

```{r}
Dgeo <- dist(EcoGenetics::ecoslot.XY(Adults.A25.ecogen))
```

### f) Mantel test

```{r}
par(mar=c(4,4,0,0))
dens <- MASS::kde2d(as.vector(Dgeo), as.vector(Dgen), n=300)
myPal <- colorRampPalette(c("white","blue","gold","orange","red"))
plot(Dgeo, Dgen, pch=20, cex=0.5,  
    xlab="Geographic Distance", ylab="Genetic Distance")
image(dens, col=transp(myPal(300), 0.7), add=TRUE)
abline(lm(Dgen ~ Dgeo))
lines(loess.smooth(Dgeo, Dgen), col="red")
```

Test for IBD with Mantel rank correlation

```{r}
IBD <- vegan::mantel(Dgen,Dgeo, method="spearman")
IBD
```

### g) Mantel correlogram

```{r}
corm <- EcoGenetics::eco.cormantel(M = Dgen, 
        XY = ecoslot.XY(Adults.A25.ecogen),  nsim = 199, latlon=FALSE, 
        alternative="less", method = "pearson", sequential=TRUE)
corm
```

### Answers 

What is the range of spatial autocorrelation in *P. vulgaris* in site A25?

- Based on a plot of genetic distance against Euclidean distance: about 15 m
- Based on where the Mantel correlogram reaches 0: < 10 m 
- Based on statistical significance tests for the Mantel correlogram: about 8.3 m


# Week 6 


### a) Load packages

```{r}
if(!require(inbreedR)) install.packages("inbreedR", repos='http://cran.us.r-project.org')
library(inbreedR)
library(dplyr)
library(ggplot2)
```

### b) Import data, extract adults
```{r}
Pulsatilla.all <- read.csv(system.file("extdata", "pulsatilla_genotypes.csv", 
                            package = "LandGenCourse"))
Adults <- Pulsatilla.all %>% filter(OffID == 0)
```

### c) Multilocus heterzygosity

Here, we're dropping the first five columns because they contain non-genetic data.

```{r}
genotypes <- inbreedR::convert_raw(Adults[,-c(1:5)])
Adults$het <- inbreedR::MLH(genotypes)
```

### d) Add population-level data

```{r}
Pop.data <- read.csv(system.file("extdata", "pulsatilla_population.csv", 
                            package = "LandGenCourse"))
Adults <- dplyr::left_join(Adults, Pop.data)
tibble::as_tibble(Adults)
```

### e) Scatterplot with line

```{r}
ggplot(Adults, aes(population.size, het)) + geom_point() +
  geom_smooth(method="lm")
```

### f) Fit linear mixed model

```{r}
mod <- lme4::lmer(het ~ population.size + (1|Population) , 
                   data = Adults, REML = TRUE)
summary(mod)
```

### g) Test fixed effect

```{r}
mod.ML <- lme4::lmer(het ~ population.size + (1|Population) , 
                   data = Adults, REML = FALSE)
car::Anova(mod.ML, type="II", test.statistic="Chisq")
```

### h) Check residual plots

```{r}
predictmeans::residplot(mod, group="Population", level=1)
```

### Answers

Conclusions from fitting the LMM:

- What was the direction of the relationship, did heterozygosity increase or decrease with census population size? - Heterozygosity decreased with census population size, which might be unexpected.
- Was the fixed effect statistically significant? If not, do you think that statistical power was sufficient?- No, the p-value was > 0.05. However, there were only seven populations, and the predictor was measured at the population level.
- Was the model valid, or was there a problem with the residual plots? - Yes, there was a problem with the residuals. 
- What would be the main issue, and what remedy could you suggest? - Residuals were not normally distributed. The response behaves like a proportion, hence fitting a generalized model with family = "binomial" might help.


# Week 7 

### a) Load packages 
    
    ```{r}
library(dplyr)
library(ggplot2)
library(nlme)
```

### b) Import data, add coordinates 

```{r}
# Dataset with variables 'flower.density' and 'mom.isolation' for each mom:
Moms <- read.csv(system.file("extdata",
                            "pulsatilla_momVariables.csv", 
                            package = "LandGenCourse"))

# Dataset with spatial coordinates of individuals:
Pulsatilla <- read.csv(system.file("extdata",
                            "pulsatilla_genotypes.csv", 
                            package = "LandGenCourse"))
Adults <- Pulsatilla %>% filter(OffID == 0)

# Combine data
Moms <- left_join(Moms, Adults[,1:5])

# Remove replicate flowers sampled from the same mother
Moms <- Moms %>% filter(OffID == 0)
```

### c) Explore data 

There are few mothers per population, however, if we can pool them among populations, we have n = 43 moms. Note that some populations have very few sampled moms.

```{r}
sum(table(Moms$Population))
Moms %>% group_by(Population) %>% summarize(n=n())
```

Both variables are strongly right-skewed:

```{r}
Histogram1 <- ggplot(Moms, aes(x=flower.density)) + geom_histogram(binwidth=5)
Histogram2 <- ggplot(Moms, aes(x=mom.isolation)) + geom_histogram(binwidth=4)

gridExtra::grid.arrange(Histogram1, Histogram2, nrow=1)
```

### d) Create scatterplots 

The double-logarithm transformation helps linearize the relationship for these data.

```{r}
myPlot <- ggplot(Moms, aes(x=mom.isolation, y=flower.density)) + 
  geom_point() 
gridExtra::grid.arrange(myPlot, myPlot + coord_trans(x="log", y="log"), nrow = 1)
```

### e) Scatterplot with line

```{r}
myPlot <- ggplot(Moms, aes(x=log(mom.isolation), y=log(flower.density))) + 
  geom_point() + geom_smooth(method="lm")
myPlot
```

### f) Fit non-spatial models

```{r}
mod.lm.REML <- gls(log(flower.density) ~ log(mom.isolation), 
                         data = Moms, method="REML")
mod.lme.REML <- lme(log(flower.density) ~ log(mom.isolation),  
                             random = ~ 1 | Population, data = Moms, method="REML")
```

### g) Plot residual variograms

The variogram for the lm-type model, without random effect, calculates distances within and between populations. The variogram is driven by comparisons between patches, whereas we are interested in spatial autocorrelation within patches. The variogram with the random effect only considers pairs within populations, hence the maximum distance lag here is around 30 m. This model clearly shows positive spatial autocorrelation over the first 5 or so meters. However, this resulted in a massive loss of pairs: the first variogram had 45 pairs per lag, the second had 6!

```{r}
semivario <- Variogram(mod.lm.REML, form= ~ X + Y, resType = "normalized")
Plot1 <- plot(semivario, smooth = TRUE)
semivario
```

```{r}
semivario <- Variogram(mod.lme.REML, form= ~ X + Y, resType = "normalized")
Plot2 <- plot(semivario, smooth = TRUE)
semivario
```
```{r}
gridExtra::grid.arrange(Plot1, Plot2, nrow = 1)
```

### h) Add correlation structure

The corGaus and corRatio models fit best. The non-spatial models fit worst.

```{r}
mod.lm.REML <- gls(log(flower.density) ~ log(mom.isolation), 
                     data = Moms, method="REML")
mod.lme.REML <- lme(log(flower.density) ~ log(mom.isolation), 
                     random = ~ 1| Population, data = Moms, method="REML")
mod.corExp<- update(mod.lme.REML, correlation = nlme::corExp(form = ~ X + Y, nugget=T))
mod.corGaus<- update(mod.lme.REML, correlation = nlme::corGaus(form = ~ X + Y, nugget=T))
mod.corSpher<- update(mod.lme.REML, correlation = nlme::corSpher(form = ~ X + Y, nugget=T))
mod.corRatio<- update(mod.lme.REML, correlation = nlme::corRatio(form = ~ X + Y, nugget=T))

MuMIn::model.sel(mod.lm.REML, mod.lme.REML, mod.corExp, mod.corGaus, mod.corSpher, mod.corRatio)
```

### i) Check residual plots 

The intercepts for populations (Random Intercepts) seem more or less normally distributed, as are the residuals. The variance is constant (Residuals vs Fitted), with a slight indication of non-linearity (tendency for positive residuals for low and high fitted values, negative for intermediate) but not troubling. The highest values tended to be underestimated (Fitted vs Observed).

```{r}
predictmeans::residplot(mod.corGaus)
```
The residual variogram is not smooth, and it still seems to increase a bit. Keep in mind that our sample size here is very small, especially as we are only considering pairs within populations. The default settings here result in only 6 pairs per distance class!

```{r}
semivario <- nlme::Variogram(mod.corGaus, form= ~ X + Y, resType = "normalized")
Plot3 <- plot(semivario, smooth = TRUE)

Fitted.variog <- nlme::Variogram(mod.corGaus)
Plot4 <- plot(Fitted.variog)

gridExtra::grid.arrange(Plot3, Plot4, nrow = 1)
```


### k) Test fixed effect

The fixed effect was statistically significant.

```{r}
car::Anova(update(mod.corGaus, method="ML"))
```

### l) Get marginal R-squared

The marginal R-squared was 45%, which is a very strong effect. This would be expected: isolated plants will have few conspecific flowers nearby.

```{r}
MuMIn::r.squaredGLMM(mod.corGaus)
```

### Answers:

- The relationship between flower density and isolation (both log-transformed) was linear, negative, very strong (45%), and statistically significant. 
- Yes: Both variables were strongly skewed, and their relationship not linear. The double logarithm transformation linearized the relationship.
- Yes: The models with spatial correlation structure performed better than the non-spatial models. I did not do a formal test for residual spatial autocorrelation though.
- Yes: This way, the variogram was only fitted to pairs within the same population (pooled across all 7 populations). 


# Week 8 

### a) Load packages

```{r}
library(dplyr)
library(gstudio)
```


### b) Import data, extract adults 

```{r}
Pulsatilla.gstudio <- gstudio::read_population(path=system.file("extdata",
                            "pulsatilla_genotypes.csv", 
                            package = "LandGenCourse"), 
                    type="column", locus.columns=c(6:19), 
                    phased=FALSE, sep=",", header=TRUE)
Adults.gstudio <- Pulsatilla.gstudio %>% filter(OffID == 0)

```

### c) Sort individuals by patch 

```{r}
# Add new variable Names:
Adults.gstudio$Names <- paste(Adults.gstudio$Population, Adults.gstudio$ID, sep=".")

# Sort alphabetically by Names:
Adults.gstudio <- dplyr::arrange(Adults.gstudio, Names)
```

### d) Calculate Euclidean distances

Note: code may differ as there are various options for identifying and specifying the correct columns to use.

```{r}
coords <- which(is.element(names(Adults.gstudio), c("X", "Y"))) 
Dgeo <- as.matrix(dist(Adults.gstudio[,coords]))
```

### e) Calculate genetic distances

```{r}
Dgen <- 1 - as.matrix(gstudio::genetic_distance(Adults.gstudio, stratum="Names", mode="dps"))
```

### f) Plot distances, calculate M.rho

Note: if plotting full matrix, the diagonal elements will be included. These may be '1' for Dgen because of the subtraction from 1 to convert to distances.

The distances mostly reflect distances between populations. The within-patch comparisons are all lumped in the very first 'column', more or less at x = 0. Note: the response variable has a limited number of values because the proportion of shared alleles, based on seven markers, can only take a limited set of values.

```{r}
plot(as.dist(Dgen) ~ as.dist(Dgeo))
cor(as.dist(Dgen), as.dist(Dgeo), method="spearman", use="complete.obs")
```

### g) Limit to within-patch pairs

```{r}
SamePatch <- outer(Adults.gstudio$Population, Adults.gstudio$Population, FUN = "==")
SamePatch[SamePatch == "FALSE"] <- NA
Dgeo.within <- SamePatch * Dgeo
```

Plot:
```{r}
plot(as.dist(Dgen) ~ as.dist(Dgeo.within))
```

Mantel rank correlation Cor.obs:

Note: this should be about 0.172. If it is lower (e.g. around 0.12), then there may be a sorting issue. Check that individuals were sorted by Names in step c).
```{r}
Cor.obs <- cor(as.dist(Dgen),as.dist(Dgeo.within), method="spearman", use="complete.obs")
Cor.obs
```

### h) Unrestricted permutation test

```{r}
Order <- c(1:nrow(Adults.gstudio))
```

Unrestricted permutation: this permutes individuals within or among patches.
```{r}
R = 499
Cor.perm.unrestricted <- rep(NA, R)
for(r in 1:R)
{
  a <- sample(Order)
  Cor.perm.unrestricted[r] <- cor(as.dist(Dgen[a,a]),as.dist(Dgeo.within),
                                  method="spearman", use="complete.obs")
}
approx.p.unrestricted <- mean(c(Cor.obs, Cor.perm.unrestricted) >= Cor.obs)
approx.p.unrestricted
```

### i) Restricted permutation test

This permutes individuals only within patches.

```{r}
R = 499
Cor.perm.restricted <- rep(NA, R)
b <- split(Order, Adults.gstudio$Population)
for(r in 1:R)
{
  a <- unlist(sapply(b, sample))
  Cor.perm.restricted[r] <- cor(as.dist(Dgen[a,a]),as.dist(Dgeo.within), 
                                method="spearman", use="complete.obs")
}
approx.p.restricted <- mean(c(Cor.obs, Cor.perm.restricted) >= Cor.obs)
approx.p.restricted
```

### k) Compare results

```{r}
boxplot(data.frame(Unrestricted = Cor.perm.unrestricted, Restricted = Cor.perm.restricted))
lines(c(0,5), c(0,0), lty=2)
```

### Answers: 

- The observed values of M.rho is not affected by the test at all, so these should not change. The p-value could change because it depends on the simulated reference distribution. In this case, it did not change because for both tests, none of the simulated values was larger or equal the observed one.
- For the unrestricted test, the distribution was centered around zero, which is what we would expect under the null hypothesis of no correlation (i.e., absense of spatial genetic structure). However, for the restricted test, the distribution was centered around 0.55, with most values positive. The explanation is not so simple, because we still have the null hypothesis of no spatial genetic structure within populations. However, some populations were small in terms of number of idividuals sampled and also patch size and thus maximum distance between individuals. Thus only a few larger patches would have contributed to larger within-patch distances, and there may be confounding by differences in diversity between patches. The restricted permutation test should take care of all these issues.
